---
title: "Lab 2 - Data wrangling"
author: "Sheila Yu"
date: "Feburary 4th, 2026"
format: html
---

```{r}
#| eval: true
#| message: false

library(tidyverse)
data <- midwest
```

# Questions

## Part 1

### Question 1

```{r}
counties_by_state <- data %>%
  count(state, name = "n_counties") %>%
  arrange(desc(n_counties))

counties_by_state

```

The code above displays calculation the number of counties in each state. Based on the calculation, Illinois (IL) has the highest number of counties, with a total of 102 counties. Wisconsin (WI) has the lowest number of counties, with a total of 72 counties.

\newpage

### Question 2

```{r}
midwest %>%
  count(county, name = "n_states") %>%
  filter(n_states == n_distinct(midwest$state))
```

Based on the result above, the county that fills in the blank of “Look at that, there is a county called \_\_\_ in each state in this dataset!” includes Crawford, Jackson and Monroe.

\newpage

### Question 3

\newpage

### Question 5

\newpage

### Question 6

\newpage

### Question 7

\newpage

### Question 8

\newpage

## Part 2

### Question 9
